---
title: "Lecture 2 - Intro to data visualization with ggplot2"
date: "05/01/2019"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}

# Don't worry about this yet
# it just tells R to show code in the output document 
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 12, 
  fig.height = 8
)

```


# Too long; didn't listen 

If you'd prefer not to follow along this lecture, here is a fun problem you can work on instead. Re-create the figure below using the `bcj_abpm.csv` file. This data contains my blood pressure over 24 hours. Note that the second column, `data`, contains my systolic blood pressure, diastolic blood pressure, heart rate, and mean arterial pressure, separated by a space. Also, I went to bed at 11pm and woke up at 7am. Good luck!

```{r, echo=FALSE}

# Load libraries
suppressPackageStartupMessages({
  # for homework problems
  library(gapminder)
  library(tidyverse)
  library(ggrepel)
  # for optional problem
  library(gganimate)
  library(gifski)
  library(lubridate)
  library(grid)
})

theme_Publication <- function(base_size=16) {
  
  require(ggthemes)
  
  (theme_foundation(base_size=base_size)+ 
      theme(plot.title = element_text(face = "bold",
                                      size = rel(1), hjust = 0.5),
            text = element_text(),
            panel.background = element_rect(colour = NA),
            plot.background = element_rect(fill = "transparent",colour = NA),
            panel.border = element_rect(colour = 'black'),
            legend.key.size = unit(3,"line"),
            legend.key.height = unit(3,"line"),
            axis.title = element_text(face = "bold",size = rel(1)),
            axis.title.y = element_text(angle=90,vjust =2),
            axis.title.x = element_text(vjust = -0.2),
            axis.text = element_text(size=rel(1)),
            axis.line = element_blank(), # element_line(colour="black"),
            axis.ticks = element_line(),
            panel.grid.major = element_blank(), #element_line(colour="#f0f0f0"),
            panel.grid.minor = element_blank(),
            legend.key = element_rect(colour = NA),
            legend.position = "bottom",
            legend.direction = "horizontal",
            legend.title = element_text(face="italic"),
            legend.text = element_text(size=rel(1)),
            plot.margin=unit(c(10,5,5,5),"mm"),
            strip.background=element_rect(colour="black",fill="#f0f0f0"),
            strip.text = element_text(face="bold")
      ))
  
}

abpm <- read_csv('bcj_abpm.csv') %>% 
  tidyr::separate(data, into = c("sbp","dbp","hr","map")) %>% 
  mutate_if(is.character, as.numeric) %>% 
  mutate(
    awake = case_when(
      time > hms('07:00:00') & time < hms('23:00:00') ~ 1, 
      TRUE ~ 0
    ),
    order = 1:n(),
  )

vline_indx <- max(abpm$order[abpm$awake==1])
ymin = min(abpm$sbp)-10
ymax = max(abpm$sbp)

readRDS('answer.RDS')

```


# The gapminder data

Let’s take a look at the data

```{r}

# Load libraries
suppressPackageStartupMessages({
  # for homework problems
  library(gapminder)
  library(tidyverse)
  library(ggrepel)
  library(magrittr)
})

# use mutate to create a new column
# the column is a log transformation of gdpPercap, i.e. income
ggdata <- gapminder %>% 
  mutate(log_income = log(gdpPercap))

```

# Two continuous variables

**Question:** Is there any relationship between life expectancy and income? (income = `gdpPercap`)

```{r}

# pass ggdata into the ggplot function using %>% 
scat_plot <- ggdata %>% 
  # let the x-axis be the gdpPercap variable
  # let the y-axis be the life expectancy variable
  ggplot(
    aes(x = gdpPercap, y = lifeExp)
  ) 

# ggplot lets you add layers however you want
# we can add points to our base plot like so
scat_plot+geom_point()

```

So, what's wrong with this plot? 

**Skew.** We can fix that using `scale_x_log10` function or just using our own derived `log_income` variable.

```{r}

# pass ggdata into the ggplot function using %>% 
scat_plot <- ggdata %>% 
  # let the x-axis be the log income variable
  # let the y-axis be the life expectancy variable
  ggplot(aes(x = log_income, y = lifeExp)) 

# ggplot lets you add layers however you want
# we can add points to our base plot like so
scat_plot+geom_point()

```

**Ambiguous labels** The x and y axes are not clearly labeled.

```{r}

scat_plot +
  geom_point()+
  # use the labs function to set axis labels
  labs(
    # set the x-axis label
    x = 'Log income',
    # set the y-axis label
    y = 'Life expectancy, years'
  )

```

**Overall relationship summary** It looks like income is positively correlated with life expectancy. However, it's tough to see the overall effect. 

```{r}

scat_plot +
  geom_point() +
  # use the labs function to set axis labels
  labs(
    # set the x-axis label
    x = 'Log income',
    # set the y-axis label
    y = 'Life expectancy, years'
  ) +
  # check out the note that ggplot gives about geom_smooth
  geom_smooth()

```

So, `gam` means generalized additive model and `s(x)` means that `ggplot` is using a spline behind the scenes to model the relationship between the `y` variable and the `x` variable. Neat, right? That's a lot of work for one line of code. But what if we didn't want a smooth line? We can tell that to `geom_smooth`:

```{r}


scat_plot +
  geom_point() +
  labs(
    # set the x-axis label
    x = 'Log income',
    # set the y-axis label
    y = 'Life expectancy, years'
  ) +
  # add a few arguments into the geom_smooth function
  # this will use a linear model instead of a gam
  geom_smooth(
    method = 'lm',
    # you could also use your own smoother
    # e.g., y ~ poly(x, degree=4) fits a fourth degree polynomial
    formula = y ~ poly(x, degree = 6),
  )

```

**Aesthetics** grey backgrounds? No thanks (but fine if that's your preference)

```{r}

scat_plot +
  geom_point(
    # this shape lets you set fill and color of points
    shape = 21,
    # note that you can set aesthetics outside of aes()
    # if you use aes(), the values are based on variables
    color = 'black',
    fill = 'grey80'
  ) +
  labs(
    # set the x-axis label
    x = 'Log income',
    # set the y-axis label
    y = 'Life expectancy, years'
  ) +
  geom_smooth(
    # use the linear model instead of generalized additive model
    method = 'lm',
    # model the y variable as a polynomial function of x, with 4 df.
    formula = y ~ poly(x, 4),
    # set the color of the smoothing line to red
    col = 'red',
    # if se = TRUE, a ribbon is plotted to show confidence limits
    # if se = FALSE, the ribbon is omitted. 
    se = FALSE
  ) + 
  # this is a general black and white theme
  theme_bw() +
  # Here we add a couple of additional theme elements
  theme(
    # let's not have grid lines
    # element_blank() allows you to remove theme elements
    panel.grid = element_blank(),
    # text needs to be read-able
    # element_text() allows you to change text options, e.g. size
    # the axis text corresponds to tick marks in the graph
    axis.text = element_text(size=16),
    # the axis title corresponds to the x- and y-axis labs.
    axis.title = element_text(size=16)
  )

```

**Sub-groups** Maybe the relationship between income and life expectancy varies across continents? We can assess this using `facet_wrap`

```{r}

scat_plot +
  geom_point(
    shape = 21,
    col = 'black',
    fill = 'grey80'
  ) +
  labs(
    # set the x-axis label
    x = 'Log income',
    # set the y-axis label
    y = 'Life expectancy, years'
  ) +
  geom_smooth(
    # use the linear model instead of generalized additive model
    method = 'lm',
    # model the y variable as a polynomial function of x, with 4 df.
    formula = y ~ poly(x, 4),
    # set the color of the smoothing line to red
    col = 'red',
    # if se = TRUE, a ribbon is plotted to show confidence limits
    # if se = FALSE, the ribbon is omitted. 
    se = FALSE
  ) + 
  # this is a general black and white theme
  theme_bw() +
  # Here we add a couple of additional theme elements
  theme(
    # let's not have grid lines
    # element_blank() allows you to remove theme elements
    panel.grid = element_blank(),
    # text needs to be read-able
    # element_text() allows you to change text options, e.g. size
    # the axis text corresponds to tick marks in the graph
    axis.text = element_text(size=16),
    # the axis title corresponds to the x- and y-axis labs.
    axis.title = element_text(size=16)
  ) +
  # the facet_wrap and facet_grid functions take a formula 
  # as an input. Formulas are special objects in R that
  # contain a ~ symbol. Note that the facet variable comes in
  # on the right hand side when we use facet_wrap with one variable.
  facet_wrap(~continent)



```

Now let's filter the data down before we plot it, and take a closer look at countries in America.

```{r}


ggdata %>% 
  dplyr::filter(continent == 'Americas') %>% 
  ggplot(aes(x = log_income, y = lifeExp)) +
  geom_point(
    shape = 21,
    col = 'black',
    fill = 'grey80'
  ) +
  labs(
    x = 'Log Income',
    y = 'Life expectancy, years'
  ) +
  geom_smooth(
    method = 'lm',
    formula = y ~ poly(x, 4),
    col = 'red',
    se = FALSE
  ) + 
  theme_bw() +
  theme(
    # let's not have grid lines
    panel.grid = element_blank(),
    # text needs to be read-able
    axis.text = element_text(size=16),
    axis.title = element_text(size=16)
  ) +
  facet_wrap(~country)

```

**Your turn!** Try looking at a different set of countries by changing the continent that you filter to. Which panels stand out? Why?

<br>
<br>

## Theme sets

So, theme elements are a lot of code, and it would be nice to just set your favorite theme once and not worry about specifying it for every plot you make. Here's how to do that

```{r}

theme_set(
  new = theme_bw() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_text(size=16),
    axis.title = element_text(size=16),
    # let's make legend text a little bigger and italic
    # this only applies to text IN the legend
    legend.text = element_text(size = 16, face = 'italic'),
    # this applies to the legend title
    legend.title = element_text(size = 16, face = 'bold.italic')
  )
)

```


# One continuous variable 

Let's take a closer look at income

```{r message=FALSE, warning=FALSE}

# hist_plot is a blank canvas 
hist_plot <- ggdata %>% 
  ggplot(aes(x=gdpPercap)) +
  geom_histogram(
    fill = 'grey80',
    col = 'black'
  )

# add a bar geom to it
hist_plot

```

Not surprising! A lot of skew in the income distribution. But what about comparing distributions in the different continents? we could facet, but that makes it hard to compare distributions between groups. Instead, we will use `ggridges`, an R package that adds a little content to `ggplot`

```{r}

# this package is needed for ridge plots
library(ggridges)

ggdata %>% 
  ggplot(
    aes(
      x = gdpPercap,
      y = continent,
      # note the use of ..density.., which is an internal
      # variable that ggplot creates for you. We can reference
      # the ..density.. variable without creating it as long
      # as the geom is compatible with densities
      height = ..density..
    )
  ) +
  geom_density_ridges(
    # this parameter sets the wiggliness of the density estimates.
    # lower values => more wiggle
    bandwidth=300
  )


```


# One categorical variable 

Let's take a look at the continents in the data. How many rows does each continent represent?

```{r}

ggdata %>% 
  # note that we are only specifying the x-variable
  ggplot(aes(x=continent))+
  # this geom intrinsically computes a statistic when it is 
  # called. The statistic that is called here is a count
  geom_bar()

```

The order doesn't seem right. What could we do to fix this?

```{r}

ggdata %>% 
  # reorder the levels of the continent variable
  # by frequencies. This will rearrange the x-axis.
  mutate(continent = fct_infreq(continent)) %>% 
  # with the levels of continent sorted by frequency,
  # the most frequently appearing variables are plotted first.
  ggplot(aes(x=continent))+
  geom_bar()


```

That looks a little better. **Question for you** Where are these counts coming from? (There is no column in the data that provides these numbers.)

```{r}

# Illustrating what ggplot is doing under the hood

ggdata %>% 
  # reorder the levels of the continent variable
  # by frequencies. This will rearrange the x-axis.
  mutate(continent = fct_infreq(continent)) %>% 
  # group these data by continent
  group_by(continent) %>% 
  # use tally to count the number of observations in each group
  tally() %>% 
  # Notice how we specify the y-variable now?
  ggplot(aes(x=continent, y = n))+
  # Now we directly specify what the bar stat is 
  # identity means we just want to plot the value of the y variable.
  geom_bar(stat = 'identity')


```

```{r}

ggdata %>% 
  # reorder the levels of the continent variable
  # by frequencies. This will rearrange the x-axis.
  mutate(continent = fct_infreq(continent)) %>% 
  # group these data by continent
  group_by(continent) %>% 
  # use summarize to count the number of observations in each group
  # AND compute the average life expectancy
  summarise(
    n = n(),
    mean_life_exp = mean(lifeExp)
  ) %>% 
  # Notice how we specify the y-variable now?
  ggplot(aes(x=continent, y = n, fill = mean_life_exp))+
  # Now we directly specify what the bar stat is 
  # identity means we just want to plot the value of the y variable.
  geom_bar(stat = 'identity')

```


# Continuous and categorical variable 

```{r}

# This package is needed for the mean_cl_normal function
suppressPackageStartupMessages(
  library(Hmisc)
)


ggdata %>% 
  ggplot(aes(x=year, y=lifeExp, col=continent))+
  # stat_summary will apply a summary function
  # to the data, using groups determined by the 
  # aesthetics of the plot. 
  stat_summary(fun.data = mean_cl_normal)


```

The problem with this plot is that the confidence limits are overlapping. We can fix this using the `position_dodge` function.

```{r}

ggdata %>% 
  ggplot(aes(x=year, y=lifeExp, col=continent))+
  stat_summary(
    fun.data = mean_cl_normal, 
    position = position_dodge(width=5)
  )

```
